Historical Soil Organic Carbon Budget

### 2020-06-26 SOCbudgetPaper/SOCbudget.tex
update method part + smaller stuff in other section and result figure

Within the last centuries the recognition of soil organic carbon (SOC)
has developed from being a major source of agricultural carbon emissions
towards a potentially large additional sink to offset other green house
gas emissions and thus mitigate climate change. Whereas the extent of
sink potential by SOC enhancement is highly debated (add refs), it is
largely agreed upon that the SOC pool itself is the biggest terrestrial
carbon pool, over topping the atmospheric and even more the biospheric
carbon pool multiple times. Even small changes in SOC drivers might lead
to substantial shifts in earth carbon cycled and influence the
atmospheric CO2 concentration (ref. permafrost melting). The specific
amount of carbon stored in the soil is uncertain though and estimates
ranging from 1500 to 2400 GtC for the first meter of the soil profile
(Bathjes, 1996).

Mapping the worlds SOC has been of great interest over the last decades
and led to an increased quality of SOC maps as well as a even better
understanding of factors driving magnitude, distribution and dynamics of
SOC pools. It is undisputed that natural properties like climatic,
biophysical and landscape characteristics play the most important role
in this regard. These factors can not be easily altered by human
intervention, leaving land cover, land use and land management changes
the most important factors for SOC dynamics till dawn of civilization .

Recent studies identified the SOC debt of anthropogenic source at around
116 GtC (Sanderman et al.), which compares to older estimates of around
60-130 GtC (Lal, 2006). Other studies have focus more closely on
spatially disaggregation SOC changes via advanced digital soil mapping
techniques (S-World; Stoorvogel 2, 2017) or better representation of
biogeochemcial processes within SOC dynamics (). Most of these studies
doing cutting edge research in their field, but seem to lack proper
representation of available information on land management.

On the other hand small-scale models (ref. Daycent, RothC, Ecosse,
C-Tool) are able to represent fairly well SOC dynamics on field-scale
driven by high quality management data on yield levels, fertilizer
inputs and various other on farming activities. Due to the lack of
proper, concise management data to feed these input demanding models on
larger extends, it seems still very complex to scale them up to a global
extent.

This study however wants to contribute filling the gap on impacts of
agricultural management like yield levels, residue recycling, manure
amendments, irrigation and tillage on SOC stocks of the worlds
croplands. It is doing so by combining work on agricultural management
data on global level with a robust but light weighted SOC model to
estimate SOC stocks and stock change factors as well as OC flow dynamics
within the agricultural system.

This is critically important since managing agricultural soil might be
one of the few setscrew for humans to \emph{naturally} combat climate
change (ref. to natural climate solutions). Other \emph{natural}
options, which might be summarized by restoring potential natural stocks
via renaturation of managed land, seem to be often limited due to
increased pressure on the land systems for feed, food and energy demand
of a growing and developing population on earth.

@@

@@

Following the tier 2 approach of the refinement of IPCC guidelines
vol.~4 (\citet{ipcc_2019_2019}), we estimate global land-use type
specific soil organic carbon (SOC) stocks for cropland and natural
vegetation on half-degree resolution from 1965 to 2010. We assume the
actual SOC state converges towards a stable steady state, that itself is
changing over time and space depending on biophysical, climatic and
agronomic conditions. Therefore we conduct the following three steps
within each yearly timestep: (1) We calculate annual land-use (sub-)type
specific steady states and decay rates for SOC stocks, (2) We account
for land conversion by transferring SOC between land-use types and (3)
We estimate SOC stocks based on the stocks of the previous time period,
the steady state stocks and the decay rate.

@@

In a simple first order kinetic approach the steady-state soil organic
carbon stocks \(SOC^{eq}\) are given by \begin{equation}
SOC^{eq}_t =\frac{C^{\textrm{in}}_t}{k_t}
\label{eq:inoutflow}
\end{equation} with \(C^{\textrm{in}}\) being carbon inputs to the soil
and \(k\) denoting the soil organic carbon decay rate. We use for our
calculations the steady-state method of the refinement of the IPCC
guidelines vol.~4 (\citet{ipcc_2019_2019}) for mineral soils, which
assume three soil carbon sub-pools (active, slow and passive) and
entangled dynamics between them. Annual carbon inflow to each sub-pool
(see @ref(sec:carboninputs)) and annual decay rates (see
@ref(sec:tier2)) of each sub-pool are still the key components to
determining steady-state SOC stocks.

BB: I would include the t, also to show that the steady-state is
time-dependent.

@@

We account for different carbon input sources depending on the land-use
type (see table @ref(tab:datasourceinputs)). Name here the inputs, and
refer to the respective sections, dont just refer to the table.
Following the IPCC methodology carbon inputs are disaggregated into
metabolic and structural components depending on their lignin and
nitrogen content (see @ref(ipcc\_2019\_2019)). For each component the
sum over all carbon input sources is allocated to the respective SOC
sub-pools via transfer coefficients. This implies that not only the
amount of carbon, but also their structural composition is determining
the effective inflow. Data sources for all considered carbon inputs as
well as for lignin and nitrogen content can be found in table
@ref(tab:datasourceinputs).

 \begin{table*}[h]
 \caption{Type and data sources for carbon inputs to different land-use types }
 \begin{tabular}{l l l l}
 \tophline
  land-use types BB: table header should be bold   & source of carbon inputs & data source & nitrogen and lignin content \\
 \middlehline
 \multirow{3}{*}{Cropland} & residues & FAOSTAT, LPJmL4 [2, sec:residues] & default values given by [2]  \\
                            & dead below ground biomass of crops & FAOSTAT, LPJmL4 [2, sec:residues] & default values given by [2] \\
                            & manure & FAOSTAT, Isabelle [2, sec:manure] & default values given by [2] \\
                            \hline
  Natural vegetation        & annual litterfall & LPJmL4 [4]& \begin{minipage}[t]{0.28\columnwidth}\raggedright\strut Nitrogen and lignin content of tree compartments used in CENTURY [4] \strut \end{minipage}\tabularnewline
 \bottomhline
 \end{tabular}
 \label{tab:datasourceinputs}
 \belowtable{}
 \end{table*}

@@

\begin{equation}
SOC_{lut,t*} = SOC_{lut,t-1} - \frac{SOC_{lut,t-1}}{A_{lut,t-1}} \cdot  AR_{lut,t} + \frac{SOC_{!lut,t-1}}{A_{!lut,t-1}} \cdot  AE_{lut,t} \qquad \forall sub, i  
\label{eq:ctransfer}
\end{equation}

@@

\begin{equation}
SOC_{t} = SOC_{t-1} + (SOC^{eq}_{t} - SOC_{t-1}) \cdot k_{t} \cdot 1\unit{a} \qquad \forall lut, sub, i.
\label{eq:steadystate}
\end{equation}

This equation can also be reformulated to a massbalance equation as
follows:

\begin{equation}
SOC_{t} = SOC_{t-1} - k_{t} \cdot 1\unit{a} + C^{\textrm{in}}  \qquad \forall lut, sub, i.
\label{eq:steadystate}
\end{equation}

@@

Additionally to the tier 2 approach of the refinement of IPCC guidelines
vol.~4 (\citet{ipcc_2019_2019}), we also estimate SOC pools using the
IPCC tier 1 approach of IPCC guidelines vol.~4 (\citet{ipcc_2006_2006})
for comparison. Here, stocks are estimated via stock change factors
given by the IPCC for the topsoil (0-30 cm) and based on a review of
measurement data. The factors differentiate different crop and
management systems reflecting different dynamics under changed in- and
outflows without explicitly tracking these. The SOC stocks as thus
calulated

\begin{equation}
SOC_{\text{target}} = \sum_{c,s,i} SOC_{\text{ref}_{c,s,i}} \cdot F_{\text{LU}_{c,s,i}} \cdot F_{\text{MG}_{c,s,i}} \cdot F_{\text{I}_{c,s,i}} \cdot A_{c,s,i}
\label{eq:tier1}
\end{equation}

\textless!- also include an equation here --\textgreater{} \textless!-
even if there are just ``copied'' out of te guidelines so to say?
--\textgreater{} \textless!- more details will follow - how deep to go?
--\textgreater{}

@@

Agricultural management data is estimated using the R library moinput
({[}cite git link{]}), which compiles country-specific FAO production
and cropland statistics (\citep{FAOSTAT}) to a comprehensive and
constistent data suite. BB: Isnt the whole calculation based on R
libraries? The data is prepared in 5 year time steps from 1965 to 2010,
which also restricts our analysis to this time span. For all the
following data, if not declared differently, we interpolate values
linearly between the time steps and hold it constant before the first
time step from 1961 to 1965. BB: too short for a spin-up, so dont use
this word. Maybe also just ignore this detail.

@@

Land-use patterns are based on the Land-Use Harmonization 2 (LUH2,
\citep{LUH2}) data set, which we aggregate from quarter degree to half
degree resolution. We disaggregate the five different cropland
subcategories (c3ann, c3per, c4ann, c4per, c3nfx) of LUH2 into our 17
crop groups, applying the relative shares for each gridcell based on the
country and year specific area harvested shares of FAOSTAT data
(\citep{FAOSTAT}) (see @ref(append:Table\_luh2fao2mag) for more details
on the crop type mapping). Land-use transitions are calculated as net
area differences of the land-use data on half-degree.

\subsubsection{Crop, Crop Residues and Pasture Production (300)}

@@

Crop residue production and management is based on a revised methodology
of (\citep{bodirsky2012}) and will be explained in key aspects again due
to its central role for soil carbon modelling. Starting from crop
production estimates of the harvested organs and their respective crop
area, we estimate above-ground (ag) and below-ground (bg) residual
biomass using yield-dependent harvest indices and shoot:root ratios. We
assume that all bg residues are recycled to the soil, whereas ag
residues can be burned or harvested for other purposes such as feeding
animals (\citep{weindl}), fuel or for material use. BB: One new thing is
here that we use a doublecropping factor to account for lower harvest
indices of multicropping vs a very high single-cropped yield

@@

A fixed share of the ag residues is assumed to be burned on field
depending on the per-capita income of the country. Following
\citep{smil1999}) we assume 25\% burn share for low-income countries
according to worldbank definitions (\(<\,1000\,\tfrac{USD}{yr}\)), 15\%
for high-income (\(>\,10000\,\tfrac{USD}{yr}\) and linearly interpolate
shares for all middle-income countries depending on their per-capita
income. Depending on the crop type 80--90\% of the residue carbon burned
on the fields are lost within the combustion process
(\citep{ipcc_2006_2006}).

@@

We compile out of our 17 crop groups, three used residue groups (straw,
high-lignin and low-lignin residues) with additional demand for other
purposes and one residues with no double use (see
@ref(append:Table\_kcr2kres)).

Residue feed demand for five different livestock groups is based on
country- and residue-group-specific feed basekts (see \citep{weindl})
taking available ag residual biomass as well as livestock productivity
into account.

We estimate a material-use share for the straw residues group of 5\% and
a fuel-share of 10\% for all used residues groups in low income
countries according to worldbank definitions
(\(<\,1000\,\tfrac{USD}{yr}\)). For high-income
(\(>\,10000\,\tfrac{USD}{yr}\) no withdrawl for material or fuel use is
assumend, leaving middle-income countries with linearly interpolate
shares depending on their per-capita income.

The remaining ag residues as well as all bg residues are assumend to be
recycled to the soil. We cut high recycling shares per hectar at the
95\%-percentile to corrected for outliers.

\paragraph{Pasture Production}

Using livestock production statistics as well as feed mix assumptions as
describted in (\citep{weindl}) we estimating country specific pasture
production. Following the same approach as for crop production we
disaggregate and calibrate half-degree pasture production pattern from
grass yields from LPJmL and pasture area and rangeland patterns ( (see
@ref(\#sec:landuse))) to derive half-degree pasture production patterns.
BB: Oh, does that mean we assume that only the pasture grows that is
being eaten? Thats not too much, is it? Maybe ask Isabelle for a better
solution (e.g.~direct pasture yields from LPJmL). And then the ``pasture
production'' is just the part being removed by grazing. Does the IPCC
specifiy something in respect to pastures?

@@

\subsubsection{Livestock Distribution and Manure Excretion (300)}

@@

To disaggregate country level FAOSTAT livestock production values to
half-degree pattern, we use the following rule based assumptions which
were inspired by the approach of \citep{gilbert} and uses feed basket
assumptions based on a revised methodology of \citet{weindl}. We
differentiate ruminant and monogastric systems, as well as extensive and
an intensive systems. For ruminants, we use the country-level shares of
gras within the feed baskets to split up pasture-fed, extensive systems
from the rather intensive crop-fed livestock. For extensive dairy and
ruminant meat production we estimate, that livestock is located in
proximity to pastures and rangelands and distribute manure excretion
proportional to pasture production of all half-degree gridcell of a
country. Intensive dairy and ruminant meat production is assumend to be
located proportinal to crop production to have short transport distances
for feed stuff.

BB: What does not become clear to me: Do the extensive cows only eat
pasture, and the intensive one only crops? Does that in the end mean we
just redistirbute grazed pasture to pastures and crop-based feed to
croplands proportional to production? Or is there more to it?

For poultry, egg and monogastric meat production we use the per-capita
income of the country to divide into intensive and extensive production
systems. For low-income countries according to worldbank definitions
(\textless1000 USD/yr), we assume extensive production systems. We
locate them according to built-up areas shares based on the idea that
these animals are held in households, subsistence or small-holder
farming systems with a high labour per animal ratio. Intensive
production is distributed within a country using the crop production
share, assuming that feed availability is the most driving factor for
livestock location.

@@

Manure production and management is based on a revised methodology of
(\citep{bodirsky2012}) and will be explained in key aspects again due to
its central role for soil carbon modelling. Based on the gridded
livestock distribution we calculate excretions by estimating the
nitrogen balance of the livestock system on the basis of comprehensive
livestock feed baskets (\citep{weindl}), assuming that all nitrogen in
protein feed intake, minus the nitrogen in the slaughter mass, is
excreted. Carbon in excreted manure is estimated by applying fixed C:N
ratios (given by \citep[(][]{ipcc_2019_2019}). Depending on the feed
system we assume manure to be handled in four different ways: All manure
orginated from pasture feed intake is excreted directly to pastures and
rangelands (pasture grazing), deducting manure collected as fuel. Manure
fuel shares are estimated using IPCC default values
(\citet{ippc_2006_2006}). Whereas for low-income countries according to
worldbank definitions (\textless1000 USD/yr), we adopt a share of 25\%
of crop residues in feed intake directly consumend and excreted on crop
fields (stubble grazing), we do not consider any stubble grazing in
high-income countries (\(>\,10000\,\tfrac{USD}{yr}\), leaving
middle-income countries with linearly interpolate shares depending on
their per-capita income. For all other feed items we assume the manure
to be stored in animal waste management systems associated to animal
houses. To estimate the carbon actually recycled to the soil, we account
for carbon losses during storage and recycling shares in different
animal waste management and grazing systems. Whereas we assume no losses
for pasture and stubble grazing, we consider that the manure collected
as fuel is not recycled. For manure stored in different animal waste
management system we compiled carbon loss rates partly depending on the
nitrogen loss rates as specified in \citep{bodirsky2012} (see
@ref(append:Table\_clossAWMS))).

@@

BB: Where does the irrigated area come from? Simple growing period
calculations together with irrigation shares of LUH2v2 are used to
estimate water effects on decay rates.

@@

Tillage data sets of {[}Vera, others{]} together with rules are used to
drive tillage effect on decay rates. \newpage

@@

\begin{figure}
\includegraphics[width=17.78in]{../OldDataNotTracked/images/maps/CShare_1965} \includegraphics[width=17.78in]{../OldDataNotTracked/images/maps/CIncrease_1965} \caption{Test}\label{fig:unnamed-chunk-6}
\end{figure}

\begin{figure}
\includegraphics[width=12cm]{../OldDataNotTracked/images/figs_draft-1} \caption{two column figure}\label{fig:unnamed-chunk-7}
\end{figure}

\begin{figure}
\includegraphics[width=12cm]{../OldDataNotTracked/images/figs_draft-2} \caption{two column figure}\label{fig:unnamed-chunk-8}
\end{figure}

\begin{figure}
\includegraphics[width=12cm]{../OldDataNotTracked/images/figs_draft-3} \caption{two column figure}\label{fig:unnamed-chunk-9}
\end{figure}

\begin{figure}
\includegraphics[width=12cm]{../OldDataNotTracked/images/RegionPlot+Valid} \caption{two column figure}\label{fig:unnamed-chunk-10}
\end{figure}

\begin{figure}
\includegraphics[width=12cm]{../OldDataNotTracked/images/maps/CShare_4ColorClimates} \caption{one column figure}\label{fig:unnamed-chunk-11}
\end{figure}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{par}\NormalTok{(}\DataTypeTok{mar =} \KeywordTok{c}\NormalTok{(}\DecValTok{4}\NormalTok{, }\DecValTok{4}\NormalTok{, }\FloatTok{0.1}\NormalTok{, }\FloatTok{0.1}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

@@

\caption{\label{tab:table}SCF compared to actual natural state}
\centering
\begin{tabular}[t]{l|r|r|r|r}
\hline
  & tropical\_moist & tropical\_dry & temperate\_dry & temperate\_moist\\
\hline
1965 & 0.25 & 0.33 & 0.55 & 0.48\\
\hline
1970 & 0.29 & 0.36 & 0.56 & 0.50\\
\hline
1975 & 0.32 & 0.38 & 0.57 & 0.51\\
\hline
1980 & 0.34 & 0.38 & 0.56 & 0.53\\
\hline
1985 & 0.36 & 0.40 & 0.60 & 0.56\\
\hline
1990 & 0.38 & 0.42 & 0.63 & 0.58\\
\hline
1995 & 0.40 & 0.45 & 0.63 & 0.59\\
\hline
2000 & 0.42 & 0.47 & 0.64 & 0.60\\
\hline
2005 & 0.45 & 0.54 & 0.69 & 0.65\\
\hline
2010 & 0.50 & 0.58 & 0.74 & 0.68\\
\hline
\end{tabular}
\end{table}

>>>>

The depletion of soil organic carbon (SOC) due agricultural management
has been a major source of carbon emissions, but also offers the
potential to turn into a large additional carbon sink and thus mitigate
climate change. Whereas the extent of the sink potential by SOC
enhancement is highly debated (add refs), it is largely agreed upon that
the SOC pool itself is the biggest terrestrial carbon pool, exceeding
the atmospheric and even more the biospheric carbon pool multiple times.
Even small changes in SOC drivers might lead to substantial shifts in
earth carbon cycled and influence the atmospheric CO2 concentration
(ref. permafrost melting). The specific amount of carbon stored in the
soil is uncertain though and estimates ranging from 1500 to 2400 GtC for
the first meter of the soil profile (Bathjes, 1996).

Mapping the worlds SOC has been investigated intensivly over the last
decades and led to an increased quality of SOC maps as well as a even
better understanding of factors driving magnitude, distribution and
dynamics of SOC pools. Natural properties like climatic, biophysical and
landscape characteristics clearly play the most important role in this
regard.

Human intervention, including land cover change and land management, has
however added a further driver to SOC change, which alters terrestrial
carbon pools in much shorter time scales and is likely the dominant
driver of SOC change on managed land today. Recent studies identified
the anthropogenic SOC debt at around 116 GtC (Sanderman et al.), which
compares to older estimates of around 60-130 GtC (Lal, 2006). Other
studies have focused more closely on spatial disaggregated SOC changes
via advanced digital soil mapping techniques (S-World; Stoorvogel 2,
2017) or better representation of biogeochemcial processes within SOC
dynamics (). Despite providing a rough estimate of the order of
magnitude of change, these studies lack a detailed consideration of land
management.

Field-scale models (ref. Daycent, RothC, Ecosse, C-Tool) are able to
capture these land mangement impacts by using detailed information on
yield levels, fertilizer inputs and various other on farming activities.
Yet, due to the lack of comprehensive global management data to feed
these input demanding models, it seems still very complex to scale them
up to a global extent.

Our study combines an spatial-explicit estimate of agricultural
management data on global level with a robust but light weighted SOC
model to estimate SOC stocks and stock change factors as well as organic
carbon flow dynamics within the agricultural system. We thereby consider
change in SOC caused by historical land cover change as well as of
different agricultural mangement practices, including residue recycling,
manure amendments, irrigation and tillage. We thereby provide the first
global, spatial explicit SOC maps that consider agricultural management.

This paper will introduce in the the first part of the method section
the basic concept of SOC dynamics. We continue in the second part with a
detailed description of the used global gridded management data on
production levels, residue recycling rates, manure amendments,
irrigation and tillage adoption. In a third, small part, we shortly
refer to the concept of stock change factors as outlined in the Tier 1
approach of the IPCC guidelines. In our result section we will focus on
the SOC dynamics on global croplands by analysing (1) the spatial
explicit SOC distribution and depletion over the last decades, (2)
compare the climate zone specfic stock changes with default stock
changes factors from the IPCC Tier 1 approach and (3) compare global
agricultrual carbon flows and stocks to contextualize the importants of
the different management aspects. Finally, we will discuss our findings
and their implications on SOC model development and conlude with an
analysis and outlook on the the opportunities for SOC management for
climate change mitigation and negative emissions.

@@

We compiled calculations as open-source R packages available under
github.com/pik-piam/mrcommons (management related functions) and
github.com/pik-piam/mrSOCbudget (soil dynamic related functions), which
are both based on github.com/pik-piam/madrat, a framework improving
reproducibility and transparency in data processing. In the following
chapter we outine the most important relationships and assumptions. See
table @ref(append:subsection2mrfunctions) for further information on
corresponding code within the R packages.

@@

Following the tier 2 approach of the refinement of IPCC guidelines
vol.~4 (\citet{ipcc_2019_2019}), we estimate soil organic carbon (SOC)
stocks for cropland and all other land represented by natural vegetation
on half-degree resolution from 1975 to 2010. We assume the actual SOC
state converges towards a stable steady state, that itself is changing
over time and space depending on biophysical, climatic and agronomic
conditions. Therefore we conduct the following three steps within each
yearly timestep: (1) We calculate annual land-use (sub-)type specific
steady states and decay rates for SOC stocks, (2) We account for land
conversion by transferring SOC between land-use types and (3) We
estimate SOC stocks based on the stocks of the previous time period, the
steady state stocks and the decay rate.

@@

In a simple first order kinetic approach the steady-state soil organic
carbon stocks \(SOC^{eq}\) are given by \begin{equation}
SOC^{eq} =\frac{C^{\textrm{in}}}{k} \qquad\forall\quad i,t
\label{eq:inoutflow}
\end{equation} with \(C^{\textrm{in}}\) being carbon inputs to the soil,
\(k\) denoting the soil organic carbon decay rate; as well as \(i\)
representing grid cells indecies and \(t\) years. We use for our
calculations the steady-state method of the refinement of the IPCC
guidelines vol.~4 (\citet{ipcc_2019_2019}) for mineral soils, which
assume three soil carbon sub-pools (active, slow and passive) and
entangled dynamics between them. Annual carbon inflow to each sub-pool
(see @ref(sec:carboninputs)) and annual decay rates (see
@ref(sec:tier2)) of each sub-pool are still the key components to
determining steady-state SOC stocks.

@@

We account for different carbon input sources depending on our two
distinguished land-use types (see table @ref(tab:datasourceinputs)); in
particular for recycled crop residues, below ground biomass of crops
(for both see @ref(sec:residues)) and recycled manure (see
\ref(sec:livst\_manure)) on croplands; and litterfall
(\ref(\citet{LPJmL4_1})) on natural vegetation.

Following the IPCC methodology carbon inputs are disaggregated into
metabolic and structural components depending on their lignin and
nitrogen content (see @ref(ipcc\_2019\_2019)). For each component the
sum over all carbon input sources is allocated to the respective SOC
sub-pools via transfer coefficients. This implies that not only the
amount of carbon, but also their structural composition is determining
the effective inflow. Data sources for all considered carbon inputs as
well as for lignin and nitrogen content can be found in table
@ref(tab:datasourceinputs).

 \begin{table*}[h]
 \caption{Type and data sources for carbon inputs to different land-use types }
 \begin{tabular}{l l l l}
 \tophline
  \textbf{land-use types}   & \textbf{source of carbon inputs} & \textbf{data source} & \textbf{nitrogen and lignin content} \\
 \middlehline
 \multirow{3}{*}{Cropland} & residues & FAOSTAT, LPJmL4 [2, sec:residues] & default values given by [2]  \\
                            & dead below ground biomass of crops & FAOSTAT, LPJmL4 [2, sec:residues] & default values given by [2] \\
                            & manure & FAOSTAT, Isabelle [2, sec:manure] & default values given by [2] \\
                            \hline
  Natural vegetation        & annual litterfall & LPJmL4 [4]& \begin{minipage}[t]{0.28\columnwidth}\raggedright\strut Nitrogen and lignin content of tree compartments used in CENTURY [4] \strut \end{minipage}\tabularnewline
 \bottomhline
 \end{tabular}
 \label{tab:datasourceinputs}
 \belowtable{}
 \end{table*}

@@

\begin{equation}
SOC_{lut,t*} = SOC_{lut,t-1} - \frac{SOC_{lut,t-1}}{A_{lut,t-1}} \cdot  AR_{lut,t} + \frac{SOC_{!lut,t-1}}{A_{!lut,t-1}} \cdot  AE_{lut,t} \qquad \forall\quad sub, i  
\label{eq:ctransfer}
\end{equation}

@@

\begin{equation}
SOC_{t} = SOC_{t-1} + (SOC^{eq}_{t} - SOC_{t-1}) \cdot k_{t} \cdot 1\unit{a} \qquad \forall\quad\quad lut, sub, i.
\label{eq:steadystate}
\end{equation}

Reformulated equation \ref(eq:steadystate), we obtain a massbalance
equation as follows

\begin{equation}
SOC_{t} = SOC_{t-1} - \underbrace{SOC_{t-1} \cdot k_{t} \cdot 1\unit{a}}_{\text{outflow}} + \overbrace{SOC^{eq}_{t} \cdot k_{t} \cdot 1\unit{a}}^{\text{input (using equation \ref(eq:inoutflow))}}  \qquad \forall\quad lut, sub, i.
\label{eq:steadystate}
\end{equation}

@@

@@

We compile country-specific FAO production and cropland statistics
(\citep{FAOSTAT}) to a comprehensive and constistent data suite. The
data is prepared in 5 year time steps from 1965 to 2010, which also
restricts our analysis together with the accounting for a spin up phase
to time span from 1975 to 2010. For all the following data, if not
declared differently, we interpolate values linearly between the time
steps.

@@

Land-use patterns are based on the Land-Use Harmonization 2 (LUH2,
\citep{LUH2}) data set, which we aggregate from quarter degree to half
degree resolution. We disaggregate the five different cropland
subcategories (c3ann, c3per, c4ann, c4per, c3nfx) of LUH2 into our 17
crop groups, applying the relative shares for each gridcell based on the
country and year specific area harvested shares of FAOSTAT data
(\citep{FAOSTAT}) (see @ref(append:Table\_luh2fao2mag) for more details
on the crop group mapping). Land-use transitions are calculated as net
area differences of the land-use data on half-degree.

\hypertarget{sec:residues}{%
\subsubsection{Crop and Crop Residues (300)}\label{sec:residues}}

@@

Crop residue production and management is based on a revised methodology
of (\citep{bodirsky2012}) and will be explained in key aspects again due
to its central role for soil carbon modelling. Starting from crop
production (\(CP\)) estimates of the harvested organs and their
respective harvested crop area (\(CA\)), we estimate above-ground
(\(AGR\)) and below-ground (\(BGR\)) residual biomass using crop group
(\(cg\)) specific harvest indices (\(HI\)) and root:shoot ratios
(\(RS\)) as follows

\begin{equation}
\begin{aligned}
AGR & = CP \cdot HI_{\textrm{slope}} + CA \cdot HI_{\textrm{intercept}}\qquad & \textrm{and} \\
BGR & = (CP + AGR) \cdot RS \qquad                                            & \forall\quad cg, i, t.
\label{eq:resbiomass}
\end{aligned}
\end{equation}

Following the IPCC guidelines, we split the harvest index into a yield
and an area dependend fraction (\citep{ipcc_2006_2006}). Note that
deviating from \citep{bodirsky2012} we use harvested instead of physical
crop area to account for increased residue biomass due to multiple
cropping and decreased amounts on fallow land. We assume that all BGR
are recycled to the soil, whereas AGR can be burned or harvested for
other purposes such as feeding animals (\citep{weindl}), fuel or for
material use.

@@

A fixed share of the AGR is assumed to be burned on field depending on
the per-capita income of the country. Following \citep{smil1999}) we
assume 25\% burn share for low-income countries according to worldbank
definitions (\(<\,1000\,\tfrac{USD}{yr}\)), 15\% for high-income
(\(>\,10000\,\tfrac{USD}{yr}\) and linearly interpolate shares for all
middle-income countries depending on their per-capita income. Depending
on the crop group 80--90\% of the residue carbon burned on the fields
are lost within the combustion process (\citep{ipcc_2006_2006}).

@@

We compile out of our 17 crop groups, three used residue groups (straw,
high-lignin and low-lignin residues) with additional demand for other
purposes and one residues with no double use (see
@ref(append:Table\_kcr2kres)). Residue feed demand for five different
livestock groups is based on country- and residue-group-specific feed
basekts (see \citep{weindl}) taking available AGR biomass as well as
livestock productivity into account. We estimate a material-use share
for the straw residues group of 5\% and a fuel-share of 10\% for all
used residues groups in low income countries according to worldbank
definitions (\(<\,1000\,\tfrac{USD}{yr}\)). For high-income
(\(>\,10000\,\tfrac{USD}{yr}\) no withdrawl for material or fuel use is
assumend, leaving middle-income countries with linearly interpolate
shares depending on their per-capita income. The remaining AGR as well
as all BGR are assumend to be recycled to the soil. We cut high
recycling shares at \(10\tfrac{\unit{tC}}{\unit{ha}}\) to corrected for
outliers.

@@

\hypertarget{sec:livst_manure}{%
\subsubsection{Livestock Distribution and Manure Excretion
(300)}\label{sec:livst_manure}}

We assume that manure is applied at its excretion place, leaving the
livestock distribution the driving factor for the spatial pattern of
manuring.

@@

To disaggregate country level FAOSTAT livestock production values to
half-degree pattern, we use the following rule based assumptions which
were inspired by the approach of \citep{gilbert} and uses feed basket
assumptions based on a revised methodology of \citet{weindl.We}
differentiate ruminant and monogastric systems, as well as extensive and
an intensive systems. For ruminants, we assume that livestock is located
at the production spot sof the feed intake. We distingush between grazed
pasture, which is converted into livestock products from extensive
systems; and all other (crop-based) feed stuff, which we consider to be
consumend in intensive systems. For poultry, egg and monogastric meat
production we use the per-capita income of the country to divide into
intensive and extensive production systems. For low-income countries
according to worldbank definitions (\textless1000 USD/yr), we assume
extensive production systems. We locate them according to built-up areas
shares based on the idea that these animals are held in households,
subsistence or small-holder farming systems with a high labour per
animal ratio. Intensive production is distributed within a country using
again crop-based feed production share, assuming that feed availability
is the most driving factor for livestock location.

@@

Manure production and management is based on a revised methodology of
(\citep{bodirsky2012}) and will be explained in key aspects again due to
its central role for soil carbon modelling. Based on the gridded
livestock distribution we calculate excretions by estimating the
nitrogen balance of the livestock system on the basis of comprehensive
livestock feed baskets (\citep{weindl}), assuming that all nitrogen in
protein feed intake, minus the nitrogen in the slaughter mass, is
excreted. Carbon in excreted manure is estimated by applying fixed C:N
ratios (given by \citep[(][]{ipcc_2019_2019}). Depending on the feed
system we assume manure to be handled in four different ways: All manure
orginated from pasture feed intake is excreted directly to pastures and
rangelands (pasture grazing), deducting manure collected as fuel. Manure
fuel shares are estimated using IPCC default values
(\citet{ippc_2006_2006}). Whereas for low-income countries according to
worldbank definitions (\textless1000 USD/yr), we adopt a share of 25\%
of crop residues in feed intake directly consumend and excreted on crop
fields (stubble grazing), we do not consider any stubble grazing in
high-income countries (\(>\,10000\,\tfrac{USD}{yr}\), leaving
middle-income countries with linearly interpolate shares depending on
their per-capita income. For all other feed items we assume the manure
to be stored in animal waste management systems associated to animal
houses. To estimate the carbon actually recycled to the soil, we account
for carbon losses during storage and recycling shares in different
animal waste management and grazing systems. Whereas we assume no losses
for pasture and stubble grazing, we consider that the manure collected
as fuel is not recycled. For manure stored in different animal waste
management system we compiled carbon loss rates partly depending on the
nitrogen loss rates as specified in \citep{bodirsky2012} (see
@ref(append:Table\_clossAWMS))). We cut high application shares at
\(10\tfrac{\unit{tC}}{\unit{ha}}\) to corrected for outliers.

@@

We use irrigation area shares to modify the water effect on
decomposition by weighting the irrigated and rainfed water factors based
on these shares. The LUH2v2 data set provides irrigated fractions for
their cropland subcategories. We apply aggregated area shares, leading
to the same impact of irrigation of all our crop groups. Moreover we
assume the irrigation effect to te present for all month of a year, that
has been marked as irrigated.

@@

To account for tillage distribution between the three tillage types
specfied by the IPCC - full tillage, reduced tillage and no tillage -,
we use the simple assumptions that all natural land and pastures are not
tilled, whereas annual crops are under full and perennials under reduced
tillage. This rules are inspired by the approach of \citep{porvolik},
but neglects the less common adoption of no tillage farming on
agricultural land.

Additionally to the tier 2 approach of the refinement of IPCC guidelines
vol.~4 (\citet{ipcc_2019_2019}) and the detailed analysis of management
data coming with it, SOC changes can be estimated using the IPCC tier 1
approach of IPCC guidelines vol.~4 (\citet{ipcc_2006_2006},
\citet{ipcc_2019_2019}). Here, stocks are calculated via stock change
factors (SCF) given by the IPCC for the topsoil (0-30 cm) and based on
observational data. SCF are differentiated by different crop and
management systems (\(m\)) reflecting different dynamics under changed
in- and outflows without explicitly tracking these. Moreover SCF vary
for different climatic zones (\(c\)) specified by the IPCC (see
@ref(append:climatemap)). The actual SOC stocks as thus calulated based
on a given reference SOC stock by

\begin{equation}
SOC^{\text{target}}_{i} = \sum_{c,m} T_{c,i} \cdot SOC^{\text{ref}} \cdot F_{\text{LU}_{c,m}} \cdot F_{\text{MG}_{c,m}} \cdot F_{\text{I}_{c,m}} \cdot A_{m,i} \qquad\forall\quad t,
\label{eq:tier1}
\end{equation}

with \(T_{c,i}\) being the translation matrix for grid cells \(i\) into
corresponding climate zones \(c\). For this analysis we us the default
SCF as comparison and constistency check for our more detailed budget
approach.

@@

@@

\caption{\label{tab:table2}SCF compared to IPCC Tier 1}
\centering
\begin{tabular}[t]{l|l|l|l|l}
\hline
Data1 & tropical\_moist & tropical\_dry & temperate\_dry & temperate\_moist\\
\hline
IPCC2006 & 0.48 & 0.58-0.64 & 0.80 & 0.69\\
\hline
IPCC2019 & 0.83 & 0.92 & 0.76-0.77 & 0.69-0.70\\
\hline
model & 0.5 & 0.58 & 0.73 & 0.67\\
\hline
\end{tabular}
\end{table}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# All defaults}
\NormalTok{knitr}\OperatorTok{::}\KeywordTok{include\_graphics}\NormalTok{(}\StringTok{"../ResultNotebooks/Output/Images/4panelfigure.png"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics[width=16cm]{../ResultNotebooks/Output/Images/4panelfigure}
\newpage
